---
title: "STAT 641 HW 1"
author: "Jack Cunningham"
format: html
editor: visual
engine: knitr
---

## Homework 1

```{r}
require(tidyverse)
require(ISLR2)
```

1\)

a.  Since n is extremely large and p is small we would expect a more flexible approach to be better than an inflexible method. The high amount of observations help alleviate the risk of following the random error too closely. The small amount of predictors allows to avoid the curse of dimensionality, it is easier to gather nearby points that share properties in order to inform future predictions.

b.  We would expect the performance of a flexible method to do worse than an inflexible method in this case. With a small amount of observations combined with the fact we have many predictors we are likely to see over fitting become a problem in the flexible method.

c.  A flexible method will do better than an inflexible method in this case, it is more capable of capturing the true shape of f. An inflexible method would introduce a significant amount of bias.

d.  A flexible method will suffer and do worse than an inflexible method in this case. The more flexible a method gets the more it tries to fit the data, in an application where the variance of error is high a flexible method will be picking up a lot of noise from the data.

2\)

a.  This is a regression problem, our response variable is CEO salary which is a quantitative value. We state we are interested in what factors impact salary thus this is an inference problem. n = 500 and p = 3.

b.  This is a classification problem, our response variable is the result of the product which is qualitative with two possibilities, success or failure. We are interested in whether our product will be a success therefore this is a prediction problem. n = 20 and p = 13.

c.  This is a regression problem, our response variable is the percentage change in USD/Euro rates which is a quantitative value. This is a prediction problem. n = 52 and p = 3.

3\)

a.  

    ```{r}
    college <- read.csv("College.csv")

    ```

b.  

    ```{r}
    rownames(college) <- college[, 1]
    college <- college[, -1]
    ```

c.  

    ```{r}
    summary(college)
    ```

```{r}
college$Private <- as.factor(college$Private)
pairs(college[,1:10])
```

```{r}
plot(college$Private,college$Outstate, col = "maroon", xlab = "Private", ylab = "Out of state tuiton")
```

```{r}
Elite <- rep("No", nrow(college))
Elite[college$Top10perc > 50] <- "Yes"
Elite <- as.factor(Elite)
college <- data.frame(college, Elite)
```

```{r}
summary(college$Elite)
```

```{r}
plot(x = college$Elite, y = college$Outstate, xlab = "Elite", ylab = "Out of state tuiton", col = "blue")
```

```{r}
hist(college$PhD, breaks = 15, col = "pink")
```

```{r}
hist(college$Grad.Rate, breaks = 20, col = "green")
```

```{r}
hist(college$Apps)
hist(college$Expend)
hist(college$perc.alumni)
hist(college$Books)
par(mfrow = c(2, 2))
```

4\.

```{r}
?Boston
```

a\) There are 506 rows and 13 columns. Each row is an observation of a Boston Suburb and each column is a measurement of the suburb (i.e. crime rate).

b\)

```{r}
pairs(Boston[,1:5])
```

A few findings we can point out:

In areas with low industry have high amounts of residential land are zoned. This makes sense, I'd imagine the majority of industry are in areas close to the city. The further you get the more space there is to set aside large lots of land to live in.

We can also see that in areas with low residential zoning there appears to be high amounts of crime, furthering the idea those areas are closer to the city.

c\)

```{r}
pairs(crim ~ lstat + tax + medv + dis, data = Boston)
```

Crime appears to be lower in areas with higher median value homes. More affluent areas seem to not have the high amount of crime that less desirable areas do. But this could be due to theses houses being on larger plots of land and naturally people being further apart.

```{r}
plot(Boston$zn, Boston$medv)
```

It is hard to tell due to the high density of suburbs having a proportion of 0 for this residential land but there doesn't appear to be a strong relationship between median value of houses in a suburb and proportion of large plots of residential land. So it seems the more affluent the area the less crime there is.

```{r}
plot(Boston$nox, Boston$crim)
```

We can see here that crime appears to be more prevalent in areas with high nitrogen oxide concentration, indicating that crime occurs more in areas where high amounts of people/industry is.

```{r}
plot(Boston$lstat, Boston$crim)
```

In areas with very few low status rates crime is low, in poorer areas crime begins to become a problem.

d\)

```{r}
plot(Boston$crim)
```

We can see that the observations from 350-490 have higher crime than the rest of the observations.

```{r}
summary(Boston$crim)
```

We can see the range of crime rate goes from nearly 0 to 89. We can also see that the vast amount of suburbs have relatively low crime rates.

```{r}
plot(Boston$tax)
```

That same range that had high crime appear to also have the highest property taxes.

```{r}
summary(Boston$tax)
```

The range of property tax rates is 187 per \$10,000 to 711 per \$10,000.

```{r}
plot(Boston$ptratio)
```

We see here that same range have the same pupil teacher ratio which is relatively high, perhaps this set of suburbs have strict rules on their pupil-teacher ratio.

There is also a small set with a rather low ratio of 13.

```{r}
summary(Boston$ptratio)
```

The range of the pupil teacher ratio is 12.60 to 22.00.

e\)

```{r}
sum(Boston$chas)
```

There are 35 suburbs who border the Charles river.

f\)

```{r}
median(Boston$ptratio)
```

The median pupil teacher ratio is 19.05.

g\)

```{r}
Boston[order(Boston$medv),]
```

We can see that the two suburbs that share the lowest median value of owner-occupied homes are index #399 and 406.

```{r}
Boston[c(399, 406),]
```

Quickly taking mean of all the other suburbs to compare values and calculating the percentile of each feature for our two standout suburbs.

```{r}
Boston|>
  slice(-c(399,406)) |> 
  summarise(across(where(is.numeric), mean))
```

```{r}
Boston |> 
  summarise(across(where(is.numeric), percent_rank)) |> 
  slice(c(399, 406))
```

We can see that the crime of these two suburbs are much higher than the average of others, in the top two percentile of suburbs.

There is no zoning for large residential land.

There is slightly more industry than average.

They do not border Charles river.

The nitrogen level is quite a bit higher than others.

The average rooms per dwelling is a lot lower than the mean.

All owner-occupied units were built before 1940.

They are closer to Boston employment centers.

They are closer to highways.

Property tax value is higher.

The pupil-teacher ratio is higher.

They have an extremely high rate of lower status individuals.

h\)

```{r}
Boston |> 
  filter(rm > 7) |> 
  summarize(n = n())
```

There are 64 rows with an average of 7 rooms or more.

```{r}
Boston |> 
  filter(rm > 8) |> 
  summarize(n = n())
```

There are 13 suburbs with an average of 8 rooms or more.

```{r}
Boston |> 
  mutate(many_rooms = rm > 8) |> 
  group_by(many_rooms) |> 
  summarise(across(where(is.numeric), mean))
```

One can investigate any particular differences they find interesting, a few of mine:

We can see the extremely low crime rate of these suburbs, the low pupil-teacher ratio, the extremely low ratio of lower-class individuals and the extremely valuable houses.
